%
% File:     dissertation.tex
% Author:   awl8049
% Revision: $Revision: 1.2 $
%
% Document based on the template for thesis and dissertations
% developed by Steven White and Malcolm Hutson.

\documentclass[12pt]{report}	% The documentclass must be ``report''.

% Dissertation package style file.
\usepackage{template/uldiss}  		

\usepackage{amsmath,amsthm,amsfonts,amscd} % Some packages to write mathematics.
\usepackage{eucal} % Euler fonts
\usepackage{verbatim} % Allows quoting source with commands.
\usepackage{listings}
\lstloadlanguages{Matlab,C++,C,Pascal}
\lstset{
         basicstyle=\footnotesize\ttfamily, 
         %numbers=left,              
         numberstyle=\tiny,          
         %stepnumber=2,              
         numbersep=5pt,              
         tabsize=2,                  
         extendedchars=true,         
         breaklines=true,            
         keywordstyle=textbf,    
         stringstyle=\ttfamily, 
         showspaces=false,       
         showtabs=false,         
         xleftmargin=17pt,
         framexleftmargin=17pt,
         framexrightmargin=5pt,
         framexbottommargin=4pt,
         %backgroundcolor=\color{lightgray},
         showstringspaces=false  
 }
\usepackage{caption}
\DeclareCaptionFont{white}{\color{white}}
\DeclareCaptionFormat{listing}{\colorbox[cmyk]{0.43, 0.35, 0.35,0.01}{\parbox{\textwidth}{\hspace{15pt}#1#2#3}}}
\captionsetup[lstlisting]{format=listing,labelfont=white,textfont=white, singlelinecheck=false, margin=0pt, font={bf,footnotesize}}
\usepackage{pdfpages}
\usepackage{comment}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.png,.jpg,.pdf}
\graphicspath{{modelgraphics/}{schedulegraphics/}}
\usepackage[caption=false,labelfont=sf,textfont=sf,captionskip=5pt]{subfig}
%\usepackage{fixltx2e}

%\usepackage{draftwatermark}		% Uncomment this line to have the
				% word, "DRAFT," as a background
				% "watermark" on all of the pages of
				% of your draft versions. When ready
				% to generate your final copy, re-comment
				% it out with a percent sign to remove
				% the word draft before you run
				% latex for the last time.
\newcommand{\equationname}{Equation}
\newcommand{\equationnames}{Equations}
\newcommand{\figurenames}{Figures}
%\newcommand{\figurename}{Figure}

% Uncomment one of the following depending upon the type of document
% you are producing
%\prospectus
\dissertation
%\masterthesis
%\masterreport

\cacscmps

%
% Basic Information 
%

\author{Adam Wade Lewis}
  % Your name how it should normally appear across the document.
  % The graduate school requires that your name always appear identically
  %   every time that it is used. To help, we recommend use \theauthor
  %   wherever your name should be printed for consistency.

\properauthor{Lewis, Adam Wade}
  % Your proper name for alphabetizing. (Used in the abstract.)
  % Last, First Middle Suffix

\title{Energy~Conservation~and~Thermal~Management in~High-Performance
  Server~Architectures}
  % The title of your thesis/dissertation. Use a tilde (~) for any
  %   spaces that should not be broken at line breaks.

\dean{David Breaux}
  % The Dean of the Graduate School

\degree{Doctor of Philosophy}
\major{Computer Science}
\graduationmonth{Spring}      
\graduationyear{2012}   
  
\previousdegrees{Bachelor of Arts, The University of the South, 1986;
  Master of Science, University of Tennessee at Chattanooga, 1989; Doctor of Philosophy, University of Louisiana at Lafayette, \thegraduationmonth \ \thegraduationyear}

% LaTeX doesn't have a good way to count words in section, use word
% counting functions in your favorite editor and insert result here
\abstractwordcount{350}
% The UL Graduate School requires that you not include the Extended
% Heading Abstract to be included in the page count.  Means that you
% cannot use the automated page count in LaTeX for this purpose.  The
% \abstractpagecount variable addresses this problem; however, you must
% manually set that value in the same manner as \abstractwordcount.
\abstractpagecount{87}
  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Enter names of the member(s) of your committee. 
% Put one name per line with the name in square brackets. 
% The name on the last line, however, must be in curly braces.
% NOTE: The first member should be your comittee chairr.
% NOTE: Maximum six members. Minimum one member (supervisor).
%
\committeemembers
         [Nian-Feng Tzeng]
         [Magdy Bayoumi]
         [Dmitri Perkins]
         {Ashok Kumar}
	
\committeememberstitle
    [Professor of Computer Engineering]
    [Professor of Computer Engineering]
    [Associate Professor of Computer Science]
    {Assistant Professor of Computer Science}

%\supervisortitle{Dr.}   

\hyphenation{FORTRAN Hy-phen-a-tion science}
%\hyphenpenalty=100000


%\singlespacing
%\oneandonehalfspacing

%\singlespacequote
%\oneandonehalfspacequote

%\topmargin 0.125in	% Adjust this value if the PostScript file output
			% of your dissertation has incorrect top and 
			% bottom margins. Print a copy of at least one
			% full page of your dissertation (not the first
			% page of a chapter) and measure the top and
			% bottom margins with a ruler. You must have
			% a top margin of 1.5" and a bottom margin of
			% at least 1.25". The page numbers must be at
			% least 1.00" from the bottom of the page.
			% If the margins are not correct, adjust this
			% value accordingly and re-compile and print again.
			%
			% The default value is 0.125" 


\makeindex              % Make the index

\begin{document}

\titlepage              % Produces the title page.
\copyrightpage          % Produces the copyright page.
\approvalpage           % Produces the approval page

%
% Dedication, epigraph, and/or acknowledgments are optional, but must
% occur here.
%
\begin{dedication}
  Dedicated to my parents, Laura Mae and Jeter Wade Lewis, who have been
  a loving and supportive presence in my life.  It has been through
  their support that this work has been possible.
\end{dedication}

 \begin{epigraph}
 Never travel without a towel.\\
 --Douglas Adams, \emph{The Hitchhiker's Guide to the Galaxy}, 1980
 \end{epigraph}

\begin{acknowledgments}		% Optional
  The author extends his thanks to the many people whose contributions
  and support have improved the overall quality of this dissertation.
  It has been an honor to work with Professor Nian-Feng Tzeng, my
  advisor and committee chair.  I am profoundly grateful to him for his
  perceptive and extremely patient mentoring over the past seven years.
  I have become a much more competent and confident scholar as result of
  his honest and supportive feedback.  I thank Professors Magdy Bayoumi,
  Dmitri Perkins, and Ashok Kumar for serving on my committee and
  providing feedback during the defense of my dissertation.

  This research was made possible by the assistance and support of my
  collaborators and co-authors.  In particular, I am grateful to Soumik
  Ghosh for his assistance in the early development of the energy and
  thermal models in this research and to Jim Simon for his assistance in
  the collection and interpretation of the results that lead to the
  development of Chaotic Attractor Predictors.  This work would not have
  been possible without the support and assistance of the members of the
  Computer Architecture and Networking research group in the Center for
  Advanced Computer Studies.  You have my thanks for your willingness to
  tolerate my quirkiness over the past seven years.

  This research was supported in part through a fellowship and funding
  provided by the Louisiana Board of Regents and through funding
  provided by the United States Department of Energy.

\end{acknowledgments}

\tableofcontents 
\listoftables    
\listoffigures   

% For really large documents, it's best to use the LaTeX include
% function to insert each chapter into the document.  If you insist, you can
% just include each chapter here...
\input{chapter-introduction}
\input{chapter-related}
\input{chapter-systemmodel}
\input{chapter-prediction}
\input{chapter-evaluation}
\input{chapter-scheduling}
\input{chapter-scheduler-evaluation}
\input{chapter-conclusions}
\bibliography{dissertation.bib}                      
% \index{Bibliography@\emph{Bibliography}}
%
% Generate the index.
%
%\printindex     % Include the index here. Comment out this line      %
%               % with a percent sign if you do not want an index .  %
 \begin{abstract} 
   Modern processors crudely manage thermal emergencies through Dynamic
   Thermal Management (DTM), where the processor monitors the die
   temperature and dynamically adjusts the processor voltage and
   frequency (DVFS) to throttle down the processor when
   necessary. However, DVFS tends to yield marked degradation in both
   application performance and system reliability.  This dissertation
   proposes a run-time model relating server energy consumption to
   its overall thermal envelope, using hardware performance counters and
   experimental measurements.  Our approach differs from prior work by
   linking system energy input to subsystem energy consumption based on
   a small set of tightly correlated parameters.  The proposed model
   takes into account processor power, bus activities, and system
   ambient temperature for real-time prediction of the power consumption
   of long running jobs.  Using the HyperTransport and QuickPath Link
   structures as case studies and through electrical measurements on
   example server subsystems, we develop a chaotic time--series
   approximation for run-time power consumption, arriving at the Chaotic
   Attractor Predictor (CAP).  With polynomial time complexity, CAP
   exhibits high prediction accuracy, with prediction errors within
   1.6\% (or 3.3\%) for servers based on the HyperTransport bus (or the
   QuickPath Links), as verified by a set of common processor
   benchmarks.

   Based on our Chaotic Attractor Predictors, we have developed and
   evaluated an effective thread scheduler for multi-core systems.
   Besides CAPs, our scheduler uses two basic principles to minimize
   server energy consumption: (1) selecting the thread with the least
   probability of causing a DTM in the subsequent time quantum, for
   execution on the next available core, and (2) migrating execution
   threads on thermally overextended cores to other cool cores via load
   balancing.  Our scheduler is evaluated in practice to assess the
   advantages resulting from incorporating thermal-awareness into the
   existing scheduler in the FreeBSD operating system.  Experiments
   reveal that our scheduler exhibits reduction in mean core on-die
   temperatures by up to 12.8$^{\circ}$C\ under PARSEC benchmarks (which
   have multi-threaded workloads) and by up to 3.3$^{\circ}$C\ under
   mixes of SPEC CPU2006 benchmarks for concurrent execution on all four
   server cores, while experiencing only 1\% to 4\% performance
   degradation.
 \end{abstract}

\begin{biography}
  Adam Wade Lewis received a Bachelor of Arts in Mathematics and
  a Bachelor of Arts in Theatre from the The University of the
  South in 1986 and a Master of Science in Computer Science from
  The University of Tennessee at Chattanooga in 1989. His background
  includes experience in the retail information technology and
  networking engineering fields as a Software Developer, Business
  Analyst, and Project Manager for NCR~Corporation throughout the
  world. He is a member of SIAM, the ACM, and the IEEE Computer Society.
 \end{biography}


\end{document}
% The following comment block is used by the different flavors of EMACS and
% the AUCTEX package to manage multiple documents.  In order for AUCTEX
% to understand you're working with multiple files, you should define
% the TeX-master variable as a file local variable that identifies your
% master document.
%
% Please do not remove.
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "dissertation.tex"
%%% TeX-PDF-mode: t
%%% TeX-source-correlate-mode: t
%%% End: 
